Guaranteeing Safety and Liveness of Unmanned Aerial Vehicle Platoons on Air Highways

### 2016-01-20 docs/AIAA 2016/plat_safety.tex
no message

% !TEX root = platooning.tex
\section{Hamilton-Jacobi Reachability  \label{sec:HJI}}
\subsection{General Framework}
Consider a differential game between two players described by the system
\begin{equation} \label{eq:dyn}
\dot{x} = f(x, u_1, u_2), \text{for almost every }t\in [-T,0]
\end{equation}

\noindent where $x\in\mathbb{R}^n$ is the system state, $u_1\in \mathcal{U}_1$ is the control of Player 1, and $u_2\in\mathcal{U}_2$ is the control of Player 2. We assume $f:\mathbb{R}^n\times \mathcal{U}_1 \times \mathcal{U}_2 \rightarrow \mathbb{R}^n$ is uniformly continuous, bounded, and Lipschitz continuous in $x$ for fixed $u_1,u_2$, and the control functions $u_1(\cdot)\in\mathbb{U}_1,u_2(\cdot)\in\mathbb{U}_2$ are drawn from the set of measurable functions. Player 2 is allowed to use nonanticipative strategies \cite{Evans84,Varaiya67} $\gamma$, defined by

\begin{equation}
\begin{aligned}
\gamma &\in \Gamma := \{\mathcal{N}: \mathbb{U}_1 \rightarrow \mathbb{U}_2 \mid  u_1(r) = \hat{u}_1(r) \\
&\text{for almost every } r\in[t,s] \Rightarrow \mathcal{N}[u_1](r) \\
&= \mathcal{N}[\hat{u}_1](r) \text{ for almost every } r\in[t,s]\}
\end{aligned}
\end{equation}

In our differential game, the goal of Player 2 is to drive the system into some target set $\mathcal{L}$, and the goal of Player 1 is to drive the system away from it. The set $\mathcal{L}$ is represented as the zero sublevel set of a bounded, Lipschitz continuous function $l:\mathbb{R}^n\rightarrow\mathbb{R}$. We call $l(\cdot)$ the \textit{implicit surface function} representing the set $\mathcal{L}=\{x\in\mathbb{R}^n \mid l(x)\le 0\}$.

Given the dynamics \eqref{eq:dyn} and the target set $\mathcal{L}$, we would like to compute the backwards reachable set, $\mathcal{V}(t)$:

\begin{equation}
\begin{aligned}
\mathcal{V}(t) &:= \{x\in\mathbb{R}^n \mid \exists \gamma\in\Gamma \text{ such that } \forall u_1(\cdot)\in\mathbb{U}_1, \\
&\exists s \in [t,0], \xi_f(s; t, x, u_1(\cdot), \gamma[u_1](\cdot)) \in \mathcal{L} \}
\end{aligned}
\end{equation}
where $\xi_f$ is the trajectory of the system satisfying initial conditions $\xi_f(t; x, t, u_1(\cdot), u_2(\cdot))=x$ and the following differential equation almost everywhere on $[-t, 0]$:
\begin{equation}
\begin{aligned}
\frac{d}{ds}&\xi_f(s; x, t, u_1(\cdot), u_2(\cdot)) \\
&= f(\xi_f(s; x, t, u_1(\cdot), u_2(\cdot)), u_1(s), u_2(s))
\end{aligned}
\end{equation}

For this paper, we use the HJ formulation in \cite{Mitchell05}, which has shown that the backwards reachable set $\mathcal{V}(t)$ can be obtained as the zero sublevel set of the viscosity solution \cite{Crandall84} $V(t,x)$ of the following terminal value Hamilton-Jacobi-Isaacs (HJI) PDE:

\begin{equation} \label{eq:HJIPDE}
\begin{aligned}
D_t &V +\min \{0, \max_{u_1\in\mathcal{U}_1} \min_{u_2\in\mathcal{U}_2} D_x V \cdot f(x,u_1,u_2) \} = 0, \\
&V(0,x) = l(x)
\end{aligned}
\end{equation}

\noindent from which we obtain $\mathcal{V}(t) = \{x\in\mathbb{R}^n \mid V(t,x)\le 0\}$. From the solution $V(t,x)$, we can also obtain the optimal controls for both players via the following:

\begin{equation} \label{eq:HJI_ctrl_syn}
\begin{aligned}
u_1^*(t,x) &= \arg \max_{u_1\in\mathcal{U}_1} \min_{u_2\in\mathcal U_2} D_x V(t,x) \cdot f(x,u_1,u_2)\\
u_2^*(t,x) &= \arg \min_{u_2\in\mathcal{U}_2} D_x V(t,x) \cdot f(x,u_1^*,u_2)
\end{aligned}
\end{equation}

In the special case where there is only one player (Player 2 does not exist), we obtain an optimal control problem for a system with dynamics

\begin{equation} \label{eq:dyn_d}
\dot{x} = f(x, u), t\in [-T,0], u\in\mathcal U.
\end{equation}

The reachable set in this case would be given by the Hamilton-Jacobi-Bellman (HJB) PDE

\begin{equation} \label{eq:HJBPDE}
\begin{aligned}
D_t V(t,x) + \min \{0, \min_{u\in\mathcal{U}} D_x V(t,x) \cdot f(x,u)\} &= 0 \\
V(0,x) = l(x)&
\end{aligned}
\end{equation}

\noindent where the optimal control is given by

\begin{equation} \label{eq:HJB_ctrl_syn}
u^*(t,x) = \arg \min_{u\in\mathcal{U}} D_x V(t,x) \cdot f(x,u)
\end{equation}

For our application, we will use a several decoupled system models and utilize the decoupled HJ formulation in \cite{Chen15}, which enables real time 4D reachable set computations and tractable 6D reachable set computations.

\section{Liveness Controllers \label{sec:liveness}}
\subsection{Merging onto a Highway \label{subsec:highway_merge}}
We model the merging of a vehicle onto an air highway as a path planning problem, where we specify a target position and velocity along the highway. Thus, a vehicle would aim to drive the system \eqref{eq:dyn} to a state $\bar{x}_H=(\bar{p}_x, \bar{v}_x, \bar{p}_y, \bar{v}_y)$, or a small range of states defined by the set

\begin{equation}
\begin{aligned}
\mathcal{L}_H = \{x: |p_x-\bar{p}_x|\le r_{p_x}, |v_x-\bar{v}_x|\le r_{v_x}, \\
|p_y - \bar{p}_y| \le r_{p_y}, |v_y - \bar{v}_y|\le r_{v_y} \}.
\end{aligned}
\end{equation}

In this reachability problem, $\mathcal{L}_H$ is the target set, represented by the zero sublevel set of the function $l_H(x)$, which specifies the terminal condition of the HJB PDE to be solved. The solution we obtain, $V_H(t,x)$, is the implicit surface function representing the reachable set $\mathcal V_H(t)$; $V_H(-T,x)\le 0$, then, specifies the reachable set $\mathcal{V}_H(T)$, the set of states from which the system can be driven to the target $\mathcal{L}_H$ within a duration of $T$. This gives the algorithm for merging onto the highway:

\begin{enumerate}
\item Move towards $\bar{x}_H$ in a straight line until $V_H(-T,x)\le 0$. This simple controller is chosen heuristically.
\item Apply the optimal control extracted from $V_H(-T,x)$ according to \eqref{eq:HJB_ctrl_syn} until $\mathcal{L}_H$ is reached.
\end{enumerate}

\subsection{Merging into a Platoon \label{subsec:platoon_merge}}
We again pose the merging of a vehicle into a platoon on an air highway as a reachability problem. Here, we would like quadrotor $Q_i$ to merge onto the highway and follow another vehicle $Q_j$ in a platoon. Thus, we would like to drive the system given by \eqref{eq:rel_dyn_aug} to a specific $\bar{x}_P = (\bar{p}_{x,r}, \bar{v}_{x,r}, \bar{p}_{y,r}, \bar{v}_{y,r})$, or a small range of relative states defined by the set

\begin{equation}
\begin{aligned}
\mathcal{L}_P = \{x: |p_{x,r}-\bar{p}_{x,r}|\le r_{p_x}, |v_{x,r}-\bar{v}_{x,r}|\le r_{v_x}, \\
|p_{y,r} - \bar{p}_{y,r}| \le r_{p_y}, |v_{y,r} - \bar{v}_{y,r}|\le r_{v_y} \}
\end{aligned}
\end{equation}

The target set $\mathcal{L}_P$ is represented by the implicit surface function $l_P(x)$, which specifies the terminal condition of the HJI PDE \eqref{eq:HJIPDE}. The zero sublevel set of the solution to \eqref{eq:HJIPDE}, $V_P(-T,x)$, gives us the set of relative states from which $Q_i$ can reach the target and join the platoon following $Q_j$ within a duration of $T$. We assume that $Q_j$ moves along the highway at constant speed, so that $u_j(t)$ = 0. The following is a suitable algorithm for a vehicle merging onto a highway and joining a platoon to follow $Q_j$:

\begin{enumerate}
\item Move towards $\bar{x}_P$ in a straight line until $V_P(-T,x)\le 0$.
\item Apply the optimal control extracted from $V_P(-T,x)$ according to \eqref{eq:HJI_ctrl_syn} until $\mathcal{L}_P$ is reached.
\end{enumerate}

\subsection{Other Quadrotor Maneuvers}
For the simpler maneuvers of traveling along a highway and following a platoon, we resort to simpler controllers described below.

\subsubsection{Traveling along a highway} \label{sec:travel_hwy}
We use a model-predictive controller (MPC) for traveling along a highway at a pre-specified speed. Here, a leader quadrotor tracks a constant-altitude path, defined as a curve $\bar{p}(s)$ parametrized by $s\in[0,1]$ in $p=(p_x, p_y)$ space (position space), while maintaining a velocity $\bar{v}(s)$ that corresponds to constant speed in the direction of the highway.

\subsubsection{Following a Platoon} \label{sec:follow_platoon}
Follower vehicles use a feedback control law tracking a nominal position and velocity in the platoon, with an additional feed-forward term given by the leader's acceleration input.

The $i$-th member of the platoon, $Q_{P_i}$, is expected to track a relative position in the platoon $r^i = (r_x^i,r_y^i)$ with respect to the leader's position $p_{P_1}$, and the leader's velocity $v_{P_1}$ at all times. The resulting control law has the form:
\begin{equation}\label{eq:follow}
u^i(t) = k_p \big[p_{P_1}(t) + r^i(t) - p^i(t) \big] + k_v\big[v_{P_1}(t) - v^i(t)\big] + u_{P_1}(t)
\end{equation}
for some $k_p,k_v>0$. A simple rule for determining $r^i(t)$ in a single-file platoon is given for $Q_{P_i}$ as:
\begin{equation}\label{eq:nominal_pos}
r^i(t) = - (i-1) b \frac{v_{P_1}}{\|v_{P_1}\|_2}
\end{equation}
where $b$ is the spacing between vehicles along the platoon. and $\frac{v_{P_1}}{\|v_{P_1}\|_2}$ is the platoon leader's direction of travel.

\section{Safety Controllers \label{sec:safety}}
\subsection{Wrapping Reachability Around Existing Controllers}
A quadrotor can use a liveness controller when it is not in any danger of collision with other quadrotors or obstacles. If the quadrotor could potentially be involved in a collision within the next short period of time, it must switch to a safety controller. In this section, we will demonstrate how HJ reachability can be used to both detect imminent danger and synthesize a controller that guarantees safety within a specified time horizon. For our safety analysis, we will use the model in \eqref{eq:rel_dyn_aug}.

We begin by defining the target set $\mathcal{L}_S$, which characterizes configurations in relative coordinates for which vehicles $Q_i,Q_j$ are considered to be in collision:

\begin{equation}
\begin{aligned}
\mathcal{L}_S = \{x: &|p_{x,r}|, |p_{y,r}|\le d \vee |v_{x,i}| \ge v_\text{max} \vee |v_{y,i}| \ge v_\text{max} \}
\end{aligned}
\end{equation}

With this definition, $Q_i$ is considered to be unsafe if $Q_i$ and $Q_j$ are within a distance $d$ in both $x$ and $y$ directions simultaneously, or if $Q_i$ has exceeded some maximum speed $v_\text{max}$ in either $x$ or $y$ direction. For illustration purposes, we choose $d=2$ meters, and $v_\text{max}= 5$ m/s.

We can now define the implicit surface function $l_S(x)$ corresponding to $\mathcal{L}_S$, and solve the HJI PDE \eqref{eq:HJIPDE} using $l_S(x)$ as the terminal condition. As before, the zero sublevel set of the solution $V_S(t,x)$ specifies the reachable set $\mathcal{V}_S(t)$, which characterizes the states in the augmented relative coordinates, as defined in \eqref{eq:rel_dyn_aug}, from which $Q_i$ \textit{cannot} avoid $\mathcal{L}_S$ for a time period of $t$, if $Q_j$ uses the worst case control. To avoid collisions, $Q_i$ must apply the safety controller according to \eqref{eq:HJI_ctrl_syn} on the boundary of the reachable set in order to avoid going into the reachable set. The following algorithm wraps our safety controller around liveness controllers:

\begin{enumerate}
\item For a specified time horizon $t$, evaluate$V_S(-t,x_i-x_j)$ for all $j\in \mathcal{Q}(i)$.

$\mathcal{Q}(i)$ is the set of quadrotors with which quadrotor $i$ checks safety against. We discuss $\mathcal{Q}(i)$ in Section \ref{subsec:safety_guarantees}.

\item Use the safety or liveness controller depending on the values $V_S(-t,x_i-x_j),j\in \mathcal{Q}(i)$:

If $\exists j\in \mathcal{Q}(i),V_S(-t,x_i-x_j)\le 0$, then $Q_i,Q_j$ are in potential conflict, and $Q_i$ must use a safety controller; otherwise $Q_i$ uses a liveness controller.
\end{enumerate}

\subsection{Platoon Safety Guarantees \label{subsec:safety_guarantees}}
Under normal operations in a single platoon, each follower quadrotor $Q_{P_i},i>1$ checks whether it is in the safety reachable set with respect to $Q_{P_{i-1}}$ and $Q_{P_{i+1}}$. So $\mathcal{Q}(i) = \{P_{i+1}, P_{i-1}\}$ for $i=P_2,\ldots,P_{N-1}$. Assuming there are no nearby quadrotors outside of the platoon, the platoon leader $Q_{P_1}$ checks safety against $Q_{P_2}$, and the platoon trailer $Q_{P_N}$ checks safety against $Q_{P_{N-1}}$. So $\mathcal{Q}(P_1)=\{P_2\}, \mathcal{Q}(P_N)=\{P_{N-1}\}$. No pair of quadrotors should be in an unsafe configuration if the liveness controllers are well-designed. Occasionally, a quadrotor $Q_k$ may behave unexpectedly due to faults, which may lead to an unsafe configuration.

With our choice of $\mathcal{Q}(i)$ and the assumption that the platoon is in a single-file formation, some quadrotor $Q_i$ would get into an unsafe configuration with $Q_k$, where $Q_k$ is likely to be the quadrotor in front or behind of $Q_i$. In this case, a ``safety breach" occurs. Our synthesis of the safety controller guarantees that between every pair of quadrotors $Q_i,Q_k$, as long as $V_S(-t,x_i- x_k)>0$, $\exists u_i$ to keep $Q_i$ from colliding with $Q_k$ for a desired time horizon $t$, despite the worst case (an adversarial) control from $Q_k$. Therefore, as long as the number of ``safety breaches" is at most one, $Q_i$ can simply use the optimal control $u_i$ to avoid collision with $Q_k$ for the time horizon of $t$. Since by assumption, vehicles in platoons are able to exit the current altitude range within a duration of $t_\text{internal}$, if we choose $t=t_\text{internal}$, the safety breach would always end before any collision can occur.

>>>>

% !TEX root = main.tex
\subsection{Safety Analysis}
Under normal operations in a single platoon, each follower quadrotor $Q_i,i=P_2,\ldots,P_{N-1}$ in a platoon checks whether it is in the safety reachable set with respect to $Q_{P_{i-1}}$ and $Q_{P_{i+1}}$. So $\mathcal{Q}(i) = \{P_{i+1}, P_{i-1}\}$ for $i=P_2,\ldots,P_{N-1}$. Assuming there are no nearby quadrotors outside of the platoon, the platoon leader $Q_{P_1}$ checks safety against $Q_{P_2}$, and the platoon trailer $Q_{P_N}$ checks safety against $Q_{P_{N-1}}$. So $\mathcal{Q}(P_1)=\{P_2\}, \mathcal{Q}(P_N)=\{P_{N-1}\}$. When all quadrotors are using liveness controllers to perform their allowed maneuvers, no pair of quadrotors should be in an unsafe configuration if the liveness controllers are well-designed. However, occasionally a quadrotor $Q_k$ may behave unexpectedly due to faults or malfunctions, in which case it may come into an unsafe configuration with another quadrotor.

With our choice of $\mathcal{Q}(i)$ and the assumption that the platoon is in a single-file formation, some quadrotor $Q_i$ would get into an unsafe configuration with $Q_k$, where $Q_k$ is likely to be the quadrotor in front or behind of $Q_i$. In this case, a ``safety breach" occurs. Our synthesis of the safety controller guarantees that between every pair of quadrotors $Q_i,Q_k$, as long as $V_S(-t,x_i- x_k)>0$, $\exists u_i$ to keep $Q_i$ from colliding with $Q_k$ for a desired time horizon $t$, despite the worst case (an adversarial) control from $Q_k$. Therefore, as long as the number of ``safety breaches" is at most one, $Q_i$ can simply use the optimal control to avoid $Q_k$ and avoid collision for the time horizon of $t$. Since quadrotors in platoons are able to exit the current altitude range within a duration of $t_\text{internal}$, if we choose $t=t_\text{internal}$, the safety breach would always end before any collision can occur.
